A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2013-04-20 docs/paper_lncs.tex
Add lncs formatted version

\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
\author{
        Russell Cohen \\
            \and
        Thomas Georgiou \\
            \and
        Pedram Razavi
}
\date{\today}
\documentclass[11pt]{article}
\usepackage[pdftex]{graphicx}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage{float}

\begin{document}
\maketitle
%\doublespace

\begin{abstract}
Here we implement simple tabulation hashing with linear probing in Python
dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
we achieved significant performance wins on synthetic benchmarks, on real
benchmarks our wins were only on the order of 4\% due to interpreter overhead.
Linear probing with sample tabulation hashing does improve Python performance
and should be looked at as a replacement scheme.  As a whole, the
extra memory bandwidth
and latency required to compute the simple tabulation hash negates some of the
performance improvement, but if they were able to be further mitigated, the new scheme
should provide more than a 4\% improvement, making it even more worthwhile.
\end{abstract}

\section{Introduction}
In this project we aim to analyze and improve the performance of Python's
dictionary or dict by changing its underlying implementation.  Python is an extremely popular general purpose computing language for several reasons:
\begin{enumerate}
 \item Clean, elegant near-pseudo code syntax leads to clear code 
 \item Widely distributed and available
 \item Large quantity of external libraries for specific features 
\end{enumerate}

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy caused by fewer tables and too much memory bandwidth used by a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted using 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

@@

Figure ~\ref{fig:randstring} shows the performance for random strings.  Both
linear probing and simple tabulation result in improved performance over the
current CPython dictionary implementation but combining both techniques results
in over a 20\% improvement.  Figure ~\ref{fig:seqstring} shows very similar
results for sequential strings.

@@

Figure ~\ref{fig:chainlength} shows the average chain length, or dictionary
cell probes per lookup for a run of the benchmark on sequential strings.
Simple tabulation hashing with CPython's current collision resolution strategy
results in the fewest number of probes.  However, despite having the fewest
number of probes, its real relative performance is much worse as seen in Figure
~\ref{fig:seqstring}.  Linear probing performs faster than its average
chain length suggests, showing the impact of looking at entries in the same
cache line first rather than jumping around.

\begin{figure}[H]
  \centering
  \includegraphics[width=4.1in]{chainlength.pdf}
  \caption{Simple tabulation hashing decreases the average chain length}
  \label{fig:chainlength}
\end{figure}

@@

We were interested to calculate if the new implementation of dictionary can handle more requests in a Python web server compared to the current CPython implementation. We chose Tornado web server for this task since most of the code is written in pure Python. An instance of Tornado server was run in a virtual machine with one assigned processor. The code of the application used is provided in the appendix.

@@

This finding as well as those by other benchmarks showed that the new
implementation gives more consistent runtime on different iterations of a same
code compared to the current CPython implementation. This consistency is a
desired behavior. We suspect that this behavior is a result of the "better" randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.

As mentioned before, another set of the benchmarks that we utilized was based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications were recoded as shown in the figures.

\begin{figure}[H]
  \centering
  \includegraphics[width=12cm]{appsbench.pdf}
  \caption{Percentage of overall improvements in some real-world Python
  applications. Nearly none of the applications performed worse. Python Pickle
  and html5lib faster around 3\% to 4\%. }
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=12cm]{slowpickle.png}
  \caption{Python Pickle module benchmark. Minimum Iteration: changed from 0.705s to 0.681s: 1.04 times faster. Average iteration: changed from 0.710s to 0.686s: 1.04 times faster. Standard deviation of new implementation 2.0383 times smaller}
\end{figure}

@@

\begin{thebibliography}{10}

\bibitem{mihai} Mihai P\v{a}tra\c{s}cu and Mikkel Thorup. The power of simple tabulation hashing. In \emph{Proc. of 43rd STOC}, 2011.

\bibitem{python} Python Software Foundation \emph{},
\texttt{http://python.org} 2012.

\bibitem{hashshoot} Nick Welch. Hash-table-shootout. 
\texttt{https://github.com/mackstann/hash-table-shootout} 2012.

\bibitem{wsgi} Nicholas Piël. Benchmark of Python WSGI Servers.

\texttt{http://nichol.as/benchmark-of-python-web-servers} 2010.

\end{thebibliography}
\appendix
\newpage
\noindent{\LARGE\bfseries Appendix}
\section{Tornado Application}
Sample Tornado 2.2.1 application used for benchmark:
 \begin{verbatim}
import tornado.httpserver
import tornado.ioloop
import tornado.wsgi

>>>>

\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{float}
% \usepackage{caption}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage{url}
\usepackage{cite}
\urldef{\mailsa}\path|{rcoh,  georgiou, prazavi}@mit.edu|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}

% a short form should be given in case it is too long for the running head
\titlerunning{A Reimplementation of Python’s dict}

% the name(s) of the author(s) follow(s) next
\author{Russell Cohen\and Thomas Georgiou\and Pedram Razavi}
%
\authorrunning{A Reimplementation of Python’s dict}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Massachusetts Institute of Technology\\
77 Massachusetts Avenue  Cambridge, MA 02139, USA\\
\mailsa\\
\url{http://www.mit.edu}}

\maketitle

\begin{abstract}
Here we implement simple tabulation hashing with linear probing in Python
dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
we achieved significant performance wins on synthetic benchmarks, on real
benchmarks our wins were only on the order of 4\% due to interpreter overhead.
Linear probing with sample tabulation hashing does improve Python performance
and should be looked at as a replacement scheme.  As a whole, the
extra memory bandwidth
and latency required to compute the simple tabulation hash negates some of the
performance improvement, but if they were able to be further mitigated, the new scheme
should provide more than a 4\% improvement, making it even more worthwhile.
\keywords{Hashing, Simple Tabulation}
\end{abstract}

\section{Introduction}
In this project we aim to analyze and improve the performance of Python's
dictionary or dict by changing its underlying implementation.  Python \cite{python} is an extremely popular general purpose computing language for several reasons:
\begin{enumerate}
 \item Clean, elegant near-pseudo code syntax leads to clear code 
 \item Widely distributed and available
 \item Large quantity of external libraries for specific features 
\end{enumerate}

@@

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Fig.~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy caused by fewer tables and too much memory bandwidth used by a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted using 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

@@

Fig.~\ref{fig:randstring} shows the performance for random strings.  Both
linear probing and simple tabulation result in improved performance over the
current CPython dictionary implementation but combining both techniques results
in over a 20\% improvement.  Fig.~\ref{fig:seqstring} shows very similar
results for sequential strings.

% \noindent%
% \begin{minipage}{\linewidth}
% \makebox[\linewidth]{%
%   \includegraphics[width=4.1in]{randstring.pdf}}
%   \captionof{figure}{Random string performance}
%   \label{fig:randstring}
% \end{minipage}

% \noindent%
% \begin{minipage}{\linewidth}
% \makebox[\linewidth]{%
%   \includegraphics[width=4.1in]{seqstring.pdf}}
%   \captionof{figure}{Sequential string performance}
%   \label{fig:seqstring}
% \end{minipage}

@@

Fig.~\ref{fig:chainlength} shows the average chain length, or dictionary
cell probes per lookup for a run of the benchmark on sequential strings.
Simple tabulation hashing with CPython's current collision resolution strategy
results in the fewest number of probes.  However, despite having the fewest
number of probes, its real relative performance is much worse as seen in Fig.~\ref{fig:seqstring}.  Linear probing performs faster than its average
chain length suggests, showing the impact of looking at entries in the same
cache line first rather than jumping around.

\begin{figure}
  \centering
  \includegraphics[width=4.1in]{chainlength.pdf}
  \caption{Simple tabulation hashing decreases the average chain length}
  \label{fig:chainlength}
\end{figure}

@@

We were interested to calculate if the new implementation of dictionary can handle more requests in a Python web server compared to the current CPython implementation. We chose Tornado web server for this task since most of the code is written in pure Python \cite{wsgi}. An instance of Tornado server was run in a virtual machine with one assigned processor. The code of the application used is provided in the appendix.

@@

This finding as well as those by other benchmarks showed that the new
implementation gives more consistent runtime on different iterations of a same
code compared to the current CPython implementation. This consistency is a
desired behavior. We suspect that this behavior is a result of the ``better'' randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.

As mentioned before, another set of the benchmarks \cite{hashshoot} that we utilized was based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications were recoded as shown in the figures.

\begin{figure}
  \centering
  \includegraphics[width=12cm]{appsbench.pdf}
  \caption{Percentage of overall improvements in some real-world Python
  applications. Nearly none of the applications performed worse. Python Pickle
  and html5lib faster around 3\% to 4\%. }
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=12cm]{slowpickle.png}
  \caption{Python Pickle module benchmark. Minimum Iteration: changed from 0.705s to 0.681s: 1.04 times faster. Average iteration: changed from 0.710s to 0.686s: 1.04 times faster. Standard deviation of new implementation 2.0383 times smaller}
\end{figure}

@@

\bibliography{references}{}
\bibliographystyle{plain}
\end{document}

\appendix
\newpage
\section*{Appendix: Tornado Application}
Sample Tornado 2.2.1 application used for benchmark:
 \begin{verbatim}
import tornado.httpserver
import tornado.ioloop
import tornado.wsgi
